### 2024-10-05 AISTAT25/aistats.tex
Updates from Overleaf
Sanjoy's final version

@@

In \emph{non-parametric active learning}, the starting point is a data
set whose labels are hidden but can be obtained individually for a
price. The goal is to label the data set, or to find a good
classifier, at low cost.

We consider a formulation in which we have an arbitrary collection of
points $X = \{x_1, \ldots, x_n\}$ from some metric space
$\X$.~\footnote{Our approach works for any topological space
  that defines a sufficiently rich set of neighborhoods. For the same of concreteness, we
  restrict ourselves to metric spaces where the neighborhoods are balls.}  We can
query the label of any of these points $x$, in which case we get a
value $y \in \{-1, +1\}$ with conditional expectation
$ \eta(x) = \E[y | x] = \pr(y=1|x) - \pr(y=-1|x) $ for some arbitrary
unknown function $\eta: \X \to [-1,1]$. The \emph{Bayes-optimal} label
for $x$, which we shall denote $g^*(x)$, is $-1$ if $\eta(x) < 0$ and
$+1$ if $\eta(x) > 0$; either label can be used if $\eta(x) = 0$.  We
wish to find an algorithm that approaches the Bayes-optimal labels for
the points in $X$ using only a small number of queries, called the
{\em query complexity} of the algorithm.

@@

Prior work on this problem~(see Section~\ref{sec:relatedwork})
requires making various assumptions, typically on the smoothness of
$\eta(x)$.  In contrast, we make no assumptions on either the set $X$
or on the conditional expectation $\eta(x)$. Our algorithm is
universal and works for any $X$ and $\eta(x)$. The {\em query
  complexity} of our algorithm depends on $X$ and $\eta$, but it does
not depend on smoothness of $\eta(x)$. Instead it depends on
properties of $\eta(x)$ close to the decision boundary where $\eta(x)$
changes sign. In particular, the algorithm is effective when $\eta(x)$
is discontinuous and hence not smooth.

@@

We begin with a large collection of \emph{balls}, $\B$, of varying
sizes (Fig.~\ref{fig:challenges}(a)). We use label queries to assign
\emph{biases} $\yh(B) \in \{-1,0,+1\}$ to balls $B \in \B$, beginning
with the largest balls. A bias of $+1$ means that the average of 
$\eta(x)$ over $s \in B$ is significantly positive; bias $-1$ means it is
significantly negative; and bias $0$ means the average $\eta$ is close
to 0. As time goes on, we get biases for progressively smaller balls,
as needed. The final label assigned to any point $x$ is based on the
biases of balls containing just $x$.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{figures/ideas.pdf}
\end{center}
\caption{(a) A collection of balls $\B$. Ball $B$ has bias $\yh(B) \in \{-1,0,+1\}$. (b) If one ball has bias $+1$ and the other $-1$, points in their intersection are targeted for active querying. (c) Biases of progressively smaller balls containing $x$ might flip sign, leading to \emph{mind changes} about the label of $x$.}
\label{fig:challenges}
\end{figure}

@@

\begin{enumerate}
\item \label{challange:wheretoquery}
  {\bf Where to query?} \\
  Initial random querying tells us the biases of the largest balls,
  which gives some information about the decision boundary. In what
  parts of $\X$ might it be helpful to know biases of smaller balls,
  warranting further label queries? Our approach is simple: if a ball
  $B$ of bias $+1$ intersects a ball $B'$ of bias $-1$ then the
  intersection $B \cap B'$ is a ``known unknown'' and is targeted for
  active querying (Fig.~\ref{fig:challenges}(b)).

@@

\item \label{challange:nosmoothness}
  {\bf No smootheness assumptions}\\
  As no assumptions of smoothness are made, the sign of $\eta(\cdot)$
  can change in an arbitrarily small ball of any given point: the
  biases of balls containing $x$, say
  $B_1 \supset B_2 \supset \cdots$, might change sign a few times. At
  first, given only the bias of $B_1$, we might think $x$ has label
  $+1$. When we later estimate the bias of $B_2$, we could change our
  mind to $-1$. And then $+1$, and so on
  (Fig.~\ref{fig:challenges}(c)).

@@

\end{enumerate}

@@

The rest of the paper is organized as
follows. Section~\ref{sec:relatedwork} describes related work.
Section~\ref{sec:algorithm} presents the active learning algorithm and
Section~\ref{sec:discrete} analyzes it in the finite population
setting, taking $X$ to be an arbitrary point-set and allowing any
$\eta$ function. We identify two \emph{critical levels} for any
$x \in X$: two scales (of ball sizes) that control how many queries
are sufficient for $x$ to be correctly labeled
(Theorem~\ref{thm:label-complexity}). We instantiate these bounds in
canonical one-dimensional settings (Theorems~\ref{thm:oned-massart}
and \ref{thm:oned-monotonic}) to get label complexities logarithmic in
$|X|$. In section~\ref{sec:experiment} we describe an experiment that supports the theoretical bound given in Theorems~\ref{thm:oned-massart}.

The supplementary material in
Section~\ref{sec:FinitePopulationProofs}, which contains the proofs of
the theorems from Section~\ref{sec:discrete}.  Finally,
Section~\ref{sec:continuous} considers the statistical setting where
$X$ is drawn from an underlying distribution $\mu$ on $\X$. We give
rates of convergence under distributional conditions
(Theorem~\ref{thm:label-complexity-specific}), and instantiate them
under common assumptions from learning theory and computational
geometry.

@@

Existing work on non-parametric active learning yields algorithms that are guaranteed to work when the underlying conditional expectation $\eta(x)$ is smooth in some way. Additional assumptions are often needed.

This work can mostly be grouped by overall principle: either (1) they
seek to obtain the Bayes-optimal labels of a few well-positioned
points, and then propagate these to the rest of the space
\cite{DNZ15,H17,ASU18} or (2) they estimate the biases (positive or
negative) of entire regions at a time \cite{DH08,M12}. In this paper,
we follow the second strategy because the only reliable and
general-purpose way to assess the bias of an individual point,
$\mbox{sign}(\eta(x))$, is to query that point repeatedly; absent
smoothness assumptions, the sign can change abruptly in an arbitrarily
small ball around $x$. On the other hand, the bias of a region
$B$---the sign of the average $\eta$ value in $B$---is easy to
determine given a reasonable number of queries chosen at random from that region.

@@

The idea of predicting with balls of multiple sizes is a central theme of \cite{BDFM19}, which predicts at a point $x$ by choosing a ball centered at $x$ that is as small as possible while having a clear bias towards one label. That work does not consider active learning, which demands considerable further innovation. For instance, to assess the uncertainty at a point $x$, we need to look at balls containing $x$, not just those centered at $x$ (Fig.~\ref{fig:challenges}(b)).

Recall that the algorithm can query any $x \in X$ and receive $-1$ or $+1$ at random,
selected according to the conditional expectation function $\eta(x) = \E[y|x]$.

@@

The goal of our analysis is to characterize conditions under which the
number of queries is significantly smaller than
$\frac{|X|}{\gamma^2}\log \frac{1}{\gamma^2}$ which is the expected
  number of number of queries made by the naive algorithm that queries
  random elements of $X$ until each $x \in X$ has been sampled
  $\frac{1}{\gamma^2}$ times.

@@

The two basic entities in our algorithm are the collection of points
$X$ and a collection of balls $\B$. Sampling is organized around a the
collection $\B$. Balls are the atomic sets on which we assess label
bias.

For any ball $B \in \B$, let $X_B = X \cap B$ be a shorthand for the
data points that lie in it.  We group balls into levels by the number
of points they contain. We put $B$ at {\it level} $\ell \geq 0$ if
\begin{equation}
\frac{|X|}{2^{\ell + 1}} \leq |X_B| < \frac{|X|}{2^\ell} .
\label{eq:sampling-level}
\end{equation}
Let $\B_\ell$ consist of all balls in $\B$ that are at level
$\ell$. Thus $\B_0, \B_1, \ldots$ is a partition of $\B$, with $\B_0$
consisting of highly-populated balls and subsequent
$\B_1, \B_2, \ldots$ consisting of successively smaller balls. We will
use $\B_{\geq \ell}$ to denote all balls at levels $\ell$ or greater,
and likewise $\B_{> \ell}$, $\B_{\leq \ell}$, and so on.

Balls in lower levels contain more points, and thus their biases
(average $\eta$ values) are easier to estimate. The main loop of the
simplified learning algorithm iterates over the levels $\ell=0,1,2,\ldots$

For any $x \in \X$, let $\B(x) \subset \B$ denote the collection of
balls that contain $x$ and can thus be used in determining $x$'s
label. We again partition these balls by sampling-level, so that
$\B_\ell(x) = \B(x) \cap \B_\ell$.

\section{A general-purpose active learning algorithm}
\label{sec:algorithm}
Our algorithm has several intricacies that might make it hard to
follow at a first read.  We therefor start with a description of a
simplified version of the algorithm (Fig~\ref{alg:simple}) that is not
completely correct but serves to explain the main components of the
detailed algorithm. The detailed algorithm is described in
subsection~\ref{sec:detailedalgorithm}.

\begin{figure*}[t]{6in}
\framebox{
\begin{minipage}[t]{6in}
\begin{center}
\vspace{.1in}
\textbf{Repeat for $\ell=0,1,2,\ldots,\log |X|-1$:}
\begin{enumerate}[leftmargin=0.6cm]
\item \label{step:Uell} {\bf Compute uncertainty region:}\\ If $\ell=0$, set $U_0 = X$; otherwise set $U_\ell = \{x \in X: \yh_{\ell-1}(x) = \,!, \ \yh_\ell(x) = \bot\}$t
\item \label{step:QS} {\bf Compute Focused Query region:}\\ Compute query region be $R=  \bigcup_{x \in U} \B_\ell(x)$
and focused query region $S=\bigcup_{B \in R} B$
\item \label{step:sample} {\bf Make queries:}\\
\textbf{Repeat until} number of queries in each ball $B \in R$ is at least $C/\gamma^2$
  \begin{enumerate}[leftmargin=0.6cm]
  \item Select a {\tt query} uniformly at random from $S$
  \end{enumerate}

\item \label{step:balls} {\bf Update Ball labels:} For each ball $B \in R$, let $k(B)$ be
  the number of queries (of any kind) made inside $B$ and let
  $\widehat{\eta}(B)$ be the average of the labels returned by
  these queries.

@@

\item \label{step:examples} {\bf Update example labels:} For each $x \in U_\ell$:
  \begin{align}
    \PL_\ell(x) &= \{s \in \{-1, +1\}: \\
               & \text{s.t. there exists } B \in B_\ell(x) \text{ s.t. } \yh(B) = s \} \\ 
    \yh_\ell(x) &= 
                  \begin{cases}
                    +1 & \text{if } \PL_\ell(x) = \{+1\} \\
                    -1 & \text{if } \PL_\ell(x) = \{-1\} \\
                    0  & \text{if } \PL_\ell(x) = \{\} \\
                    !  & \text{if } \PL_\ell(x) = \{-1,+1\}
                  \end{cases}
                  \label{eq:provisional-label}
  \end{align}
\end{enumerate}
\end{center}

@@

We now describe the simplified algorithm here referred to simply as
the algorithm. The algorithm loops over increasing levels.  At each
level it identifies the region of ``known unknowns'' which approximates
the boundary. The algorithm then makes queries in the vicinity of the
boundary to refine it and proceeds to the next finer level.

What follows is a detailed explanation of  Figure~\ref{alg:simple}.
The outer loop of the algorithm iterates over the
levels, starting with $\ell=0$ that corresponds to balls that contain
at least half of the points in $X$, and ending with $\ell=\log |X|-1$,
corresponding to balls that contain a single point. Initially, all balls are labeled $\bot$, which indicates that there are not enough queries inside the ball to determine it's label.

The loop consists of five steps. The first computes the {\em uncertainty
region}, the second computes the {\em focused query region}, the third defines the
sampling process, the fourth and fifth update the labels of the balls
and the points respectively.

In step~\ref{step:Uell} the {\em uncertainty region} $U_\ell$ is
calculated. Intuitively $U_l$ is the set of points whose label
according to level $\ell-1$ is ! or ``known unknown'' and the label
according to level $\ell$ is $\bot$ or ``not enough information''.
These are the points for which we need better estimates.

In step~\ref{step:QS} the algorithm identifies the balls that needs to
be queried in order to get more information about $U_\ell$. These are
the balls in $\B_l$ that contain at least one point of $U_\ell$.  The
region $S$ is the union of these balls and defines the set
from which queries are selected in order to reduce the known unknowns.

@@

Step ~\ref{step:balls} associates a label with each ball in the focus region. The
label is one of $\{-1,+1,0\}$.  The labels $+1$ and $-1$ correspond to
a confident statement that $\eta(B)>0$ and $\eta(B)<0$
respectively. Finally, the label $0$ indicates that the sign of $B$
is uncertain given the answers to the queries.

In step~\ref{step:examples} these labels are propagated to the
examples.  This is done in two sub-steps. First the {\em Possible
  Labels} for $x$ denoted $\PL_\ell(x)$ are calculated. $\PL_\ell(x)$
contains $+1$ if there is a ball containing $x$ whose label is $+1$,
symmetrically for $-1$. The label $\yh_\ell(x)$ is determined by the
possible labels set $\PL_\ell(x)$ as described in
Equation~\ref{eq:provisional-label}. The most significant label is ``!'', which
indicates balls with conflicting polarity containing $x$. Points
labeled ``!'' determine the uncertainty region and the focused query
region in the following level and the process repeats.

@@

The simplified algorithm does addresses the first challange of ``where
to query'' (see challanges in the introduction). It does not address
the other two challanges, which the detailed algorithm does address.

\begin{enumerate}
\item {\bf No smoothness assumptions:}\\
  As described in the introduction, one can \emph{never} be sure of having
  correctly determined the label of $x$. Thus in addition to focused
  (active) querying, we also do background sampling of the entire
  space to pick up on possible mind changes. For simplicity, we make
  one background query per focused query.

@@

\item {\bf Managing Sampling:}\\
  As described in the introduction, we would like to use queries that
  fall in the intersection of two balls for estimating the bias in
  both balls. However, selecting examples IID at random from the balls
  can create an undesired dependence between the estimates. This is solved using Poisson sampling, as described in section~\ref{sec:poisson}.
\end{enumerate}

@@

Once these queries are made, bias estimates $\widehat{\eta}(B)$ are
updated using the Poisson method,\\
$\yh(B), \PL_\ell(x), \yh_\ell(x), U_\ell$ are updated
accordingly  and the next iteration begins.

@@

The uncertainty region at level $\ell$ consists of the known-unknowns
from level $\ell-1$, that is, points $x \in X$ with
$\yh_{\ell-1}(x) = \, !$. This region is revealed to us piecemeal, a
few points at a time. To get good label complexity we need to actively
query the part of it we know.

We manage this through \emph{Poisson
  sampling}~\cite{ghosh2002sampling} (Fig.~\ref{alg:sampling}) which
works as follows. Each instance $x \in X$ is assocciated with a random
draw from the uniform distribution over $[0,1]$ denoted $T_x$.
Pooints are sampled in the order of increasing $T_x$ under some
conditions described below.  Note that restricting the sampled points
to be from set $A$ and having $T_x \leq \tau$ is the same, in
expectation to sampling each point in $A$ independently at random with
probability $\tau$.

To infer the bias of a ball $B$, we need labels for
$k = \tilde{O}(1/\gamma^2)$ random points in it. Balls at level $\ell$
have roughly $n/2^{\ell+1}$ points, so we need about $2^{\ell+1} k/n$
fraction of the ball to be queried. Instead, we query every point in
the ball with probability $\tau_\ell = \min(2^{\ell+2}k/n, 1)$,
\emph{independently}. Specifically, we query points $x$ in the ball
with $T_x \leq \tau_\ell$, where each $x \in X$ is assigned a
uniform-random value $T_x \in [0,1]$ at the outset of the algorithm.

Once a point is queried, it is never queried again.

@@

Initialize uncertainty regions at all levels: $U_0 = X$ and $U_\ell = \emptyset$ for $\ell \geq 1$

>>>>

\usepackage{aistats2025}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{times}
\usepackage{amssymb}
\usepackage{graphicx}

@@

In \emph{nonparametric active learning}, the starting point is a data
set whose labels are hidden but can be obtained individually for a
price. The goal is to label the data set, or to find a corresponding
classifier, at low cost.

We consider a formulation in which we have an arbitrary collection of
points $X = \{x_1, \ldots, x_n\}$ from some metric space
$\X$.~\footnote{Our approach works for any topological space
  that defines a sufficiently rich set of neighborhoods. For concreteness, we
  focus on metric spaces and take the neighborhoods to be balls.}  We can
query the label of any of these points $x$, in which case we get a
value $y \in \{-1, +1\}$ with conditional expectation
$ \eta(x) = \E[y | x] = \pr(y=1|x) - \pr(y=-1|x) $ for some arbitrary
unknown function $\eta: \X \to [-1,1]$. The \emph{Bayes-optimal} label
for $x$, which we shall denote $g^*(x)$, is $-1$ if $\eta(x) < 0$ and
$+1$ if $\eta(x) > 0$; either label can be used if $\eta(x) = 0$.  We
wish to find an algorithm that approaches the Bayes-optimal labels for
the points in $X$ using only a small number of queries, called the
{\em query complexity} of the algorithm.

@@

Prior work on this problem~(see Section~\ref{sec:relatedwork})
requires making various assumptions, typically on the smoothness of
$\eta(x)$.  In contrast, we make no assumptions on either the set $X$
or on the conditional expectation $\eta(x)$. Our algorithm is
universal and works for any $X$ and $\eta(x)$. Our {\em query
  complexity} bound depends on $X$ and $\eta$, but has no explicit dependence on the smoothness of $\eta(x)$. Instead it depends on
properties of $\eta(x)$ close to the decision boundary where $\eta(x)$ changes sign. In particular, the algorithm is effective even when $\eta(x)$
is discontinuous.

@@

We begin with a large collection of \emph{balls}, $\B$, of varying
sizes (Fig.~\ref{fig:challenges}(a)). We use label queries to assign
\emph{biases} $\yh(B) \in \{-1,0,+1\}$ to balls $B \in \B$, beginning
with the largest balls. A bias of $+1$ means that the average of 
$\eta(x)$ over $x \in B$ is significantly positive; bias $-1$ means it is
significantly negative; and bias $0$ means the average $\eta$ is close
to 0. As time goes on, we get biases for progressively smaller balls,
as needed. The final label assigned to any point $x$ is based on the
biases of balls containing $x$.

\begin{figure}
\begin{center}
\includegraphics[width=3.25in]{figures/ideas.pdf}
\end{center}
\caption{(a) A collection of balls $\B$. Ball $B$ has bias $\yh(B) \in \{-1,0,+1\}$. (b) If one ball has bias $+1$ and the other $-1$, points in their intersection are targeted for active querying. (c) Biases of progressively smaller balls containing $x$ might flip sign, leading to \emph{mind changes} about the label of $x$.}
\label{fig:challenges}
\end{figure}

@@

\paragraph{1. Where to query?}\label{challange:wheretoquery}

  Initial random querying tells us the biases of the largest balls,
  which gives some information about the decision boundary. In what
  parts of $\X$ might it be helpful to know biases of smaller balls,
  warranting further label queries? Our approach is simple: if a ball
  $B$ of bias $+1$ intersects a ball $B'$ of bias $-1$ then the
  intersection $B \cap B'$ is a ``known unknown'' and is targeted for
  active querying (Fig.~\ref{fig:challenges}(b)).

\paragraph{2. No smoothness assumptions} \label{challange:nosmoothness}

  As no assumptions of smoothness are made, the sign of $\eta(\cdot)$
  can change in an arbitrarily small neighborhood of any given point: the
  biases of balls containing $x$, say
  $B_1 \supset B_2 \supset \cdots$, might change sign a few times. At
  first, given only the bias of $B_1$, we might think $x$ has label
  $+1$. When we later estimate the bias of $B_2$, we could change our
  mind to $-1$. And then $+1$, and so on
  (Fig.~\ref{fig:challenges}(c)).

\paragraph{3. Managing sampling} \label{challange:managingsampling}

  The third challenge is \emph{managing sampling} of overlapping
  balls. We are interested in detecting points, and thus balls, whose
  $\eta$-values are either $> \gamma$ or $<-\gamma$. This suggests
  querying $k \approx 1/\gamma^2$ points at random from each
  ball. Now, suppose we have queried this many points from ball $B$
  and later want to query points from a different ball $B'$ that
  overlaps $B$. How can we reuse queries already made in $B \cap B'$?

\iffalse
\begin{enumerate}
\item \label{challange:wheretoquery}
  {\bf Where to query?} \\
  Initial random querying tells us the biases of the largest balls,
  which gives some information about the decision boundary. In what
  parts of $\X$ might it be helpful to know biases of smaller balls,
  warranting further label queries? Our approach is simple: if a ball
  $B$ of bias $+1$ intersects a ball $B'$ of bias $-1$ then the
  intersection $B \cap B'$ is a ``known unknown'' and is targeted for
  active querying (Fig.~\ref{fig:challenges}(b)).

@@

\item \label{challange:nosmoothness}
  {\bf No smoothness assumptions}\\
  As no assumptions of smoothness are made, the sign of $\eta(\cdot)$
  can change in an arbitrarily small ball of any given point: the
  biases of balls containing $x$, say
  $B_1 \supset B_2 \supset \cdots$, might change sign a few times. At
  first, given only the bias of $B_1$, we might think $x$ has label
  $+1$. When we later estimate the bias of $B_2$, we could change our
  mind to $-1$. And then $+1$, and so on
  (Fig.~\ref{fig:challenges}(c)).

@@

\end{enumerate}
\fi

@@

The rest of the paper is organized as
follows. Section~\ref{sec:relatedwork} describes related work.
Section~\ref{sec:algorithm} presents the active learning algorithm and
Section~\ref{sec:discrete} analyzes it in the finite population
setting, taking $X$ to be an arbitrary point-set and allowing any
$\eta$ function. We identify two \emph{critical levels} for any
$x \in X$: two scales (of ball sizes) that control how many queries
are sufficient for $x$ to be correctly labeled
(Theorem~\ref{thm:label-complexity}). We instantiate these bounds in
canonical one-dimensional settings (Theorems~\ref{thm:oned-massart}
and \ref{thm:oned-monotonic}) to get label complexities logarithmic in
$|X|$. In section~\ref{sec:experiment} we describe an experiment that supports the theoretical bound given in Theorem~\ref{thm:oned-massart}.

The supplementary material in
Section~\ref{sec:FinitePopulationProofs} contains the proofs of
the theorems from Section~\ref{sec:discrete}.  Finally,
Section~\ref{sec:continuous} considers the statistical setting where
$X$ is drawn from an underlying distribution $\mu$ on $\X$. We give
rates of convergence under distributional conditions
(Theorem~\ref{thm:label-complexity-specific}), and instantiate them
under common assumptions from learning theory and computational
geometry.

@@

Existing work on nonparametric active learning yields algorithms that are guaranteed to work when the underlying conditional expectation $\eta(x)$ is smooth in some way. Additional assumptions are often needed.

This work can mostly be grouped by overall principle: either (1) they
seek to obtain the Bayes-optimal labels of a few well-positioned
points, and then propagate these to the rest of the space
\cite{DNZ15,H17,ASU18} or (2) they estimate the biases (positive or
negative) of entire regions at a time \cite{DH08,M12}. In this paper,
we follow the second strategy because the bias of a region
$B$---the sign of the average $\eta$ value in $B$---is easy to
determine given a reasonable number of queries chosen at random from that region. 
On the other hand, the only reliable and
general-purpose way to assess the bias of an individual point,
$\mbox{sign}(\eta(x))$, is to query that point repeatedly; absent
smoothness assumptions, the sign can change abruptly in an arbitrarily
small ball around $x$.

@@

The idea of predicting with balls of multiple sizes is a central theme of \cite{BDFM19}, which predicts at a point $x$ by choosing a ball centered at $x$ that is as small as possible while having a clear bias towards one label. That work does not consider active learning, which demands further innovation. For instance, to assess the uncertainty at a point $x$, we look at balls \emph{containing} $x$, not just \emph{centered} at $x$ (Fig.~\ref{fig:challenges}(b)).

\iffalse
Recall that the algorithm can query any $x \in X$ and receive $-1$ or $+1$ at random,
selected according to the conditional expectation function $\eta(x) = \E[y|x]$.

@@

The goal of our analysis is to characterize conditions under which the
number of queries is significantly smaller than
$\frac{|X|}{\gamma^2}\log \frac{1}{\gamma^2}$ which is the expected
  number of number of queries made by the naive algorithm that queries
  random elements of $X$ until each $x \in X$ has been sampled
  $\frac{1}{\gamma^2}$ times.
\fi

@@

The two basic entities in our algorithm are the collection of points
$X$ and a collection of balls $\B$. Sampling is organized around $\B$:
these balls are the atomic sets on which we assess label
bias.

For any ball $B \in \B$, let $X_B = X \cap B$ be a shorthand for the
data points lying in it.  We group balls into levels by the number
of points they contain. We put $B$ at {\it level} $\ell \geq 0$ if
\begin{equation}
\frac{|X|}{2^{\ell + 1}} \leq |X_B| < \frac{|X|}{2^\ell} .
\label{eq:sampling-level}
\end{equation}
Let $\B_\ell$ consist of all balls in $\B$ at level
$\ell$. Thus $\B_0, \B_1, \ldots$ is a partition of $\B$, with $\B_0$
consisting of highly-populated balls and subsequent
$\B_1, \B_2, \ldots$ consisting of successively smaller balls. We will
use $\B_{\geq \ell}$ to denote all balls at levels $\ell$ or greater,
and likewise $\B_{> \ell}$, $\B_{\leq \ell}$, and so on.

Balls in lower levels contain more points, and thus their biases
(average $\eta$ values) are easier to estimate. The overall querying
strategy begins with balls at the lowest levels.
%main loop of the
%simplified learning algorithm iterates over the levels $\ell=0,1,2,\ldots$

For any point $x \in \X$, let $\B(x) \subset \B$ denote the collection of
balls that contain $x$ and can thus be used in determining $x$'s
label. We again partition these balls by sampling-level, so that
$\B_\ell(x) = \B(x) \cap \B_\ell$.

\section{A general-purpose active learning algorithm}
\label{sec:algorithm}
Our algorithm has several intricacies that might make it hard to
follow at a first read.  We therefore start with a
simplified version of the algorithm (Fig~\ref{alg:simple}) that serves
to explain the main ideas but is missing some details.  
The detailed algorithm is described in
Section~\ref{sec:detailedalgorithm}.

\begin{figure*}[t]
\framebox{
\begin{minipage}[t]{6.5in}
%\begin{center}
\vspace{.1in}
\noindent\textbf{Repeat for $\ell=0,1,2,\ldots,$:}
\begin{enumerate}[leftmargin=0.6cm]
\item \label{step:Uell} {\bf Compute uncertainty region:}\\ If $\ell=0$, set $U_0 = X$; otherwise set $U_\ell = \{x \in X: \yh_{\ell-1}(x) = \,!\}$ %, \ \yh_\ell(x) = \bot\}$
\item \label{step:QS} {\bf Compute focused query region:}\\ 
The balls to be queried are $R=  \bigcup_{x \in U} \B_\ell(x)$
and the focused query region is their union, $S=\bigcup_{B \in R} B$
\item \label{step:sample} {\bf Make queries:}\\
\textbf{Repeat until} number of queries in each ball $B \in R$ is at least $C/\gamma^2$
  \begin{itemize}[leftmargin=0.6cm]
  \item Select a {\tt query} point uniformly at random from $S$
  \end{itemize}

\item \label{step:balls} {\bf Update ball labels:} For each ball $B \in R$, let 
    %$k(B)$ be the number of queries (of any kind) made inside $B$ and let
  $\widehat{\eta}(B)$ be the average of the labels returned by
  queries in $B$.

@@

\item \label{step:examples} {\bf Update example labels:} For each $x \in U_\ell$:
  \begin{align}
    \PL_\ell(x) &= \{s \in \{-1, +1\}: \\
               & \text{s.t. there exists } B \in B_\ell(x) \text{ s.t. } \yh(B) = s \} \\ 
    \yh_\ell(x) &= 
                  \begin{cases}
                    +1 & \text{if } \PL_\ell(x) = \{+1\} \\
                    -1 & \text{if } \PL_\ell(x) = \{-1\} \\
                    0  & \text{if } \PL_\ell(x) = \{\} \\
                    !  & \text{if } \PL_\ell(x) = \{-1,+1\}
                  \end{cases}
                  \label{eq:provisional-label}
  \end{align}
\end{enumerate}
%\end{center}

@@

We now describe the simplified algorithm, here referred to simply as
the algorithm. The algorithm loops over increasing levels.  At each
level it identifies the region of ``known unknowns'' which approximates
the boundary. The algorithm then makes queries in the vicinity of the
boundary to refine it and proceeds to the next finer level.

What follows is a detailed explanation of  Figure~\ref{alg:simple}.
The outer loop of the algorithm iterates over
levels, starting with $\ell=0$ that corresponds to balls with at least half of the points in $X$, then moving onto $\ell=1$ which has balls with between $1/4$ and $1/2$ of the points in $X$, and so on.
%and ending with $\ell=\log |X|-1$,
%corresponding to balls that contain a single point. %Initially, all balls are labeled $\bot$, which indicates that there are not enough queries inside the ball to determine it's label.

The loop consists of five steps. The first computes the {\em uncertainty
region}, the second computes the {\em focused query region}, the third defines the
sampling process, the fourth and fifth update the labels of balls
and points respectively.

In step~\ref{step:Uell} the {\em uncertainty region} $U_\ell$ is
calculated. Intuitively $U_l$ is the set of points whose label
according to level $\ell-1$ is ! or ``known unknown''.
%and the label
%according to level $\ell$ is $\bot$ or ``not enough information''.
These are the points for which we need better estimates.

In step~\ref{step:QS} the algorithm identifies balls that need to
be queried to get more information about $U_\ell$. These are
the balls in $\B_\ell$ that contain at least one point of $U_\ell$. 
Region $S$ is the union of these balls. It is the set
from which queries are selected in order to reduce the known-unknowns.

@@

Step ~\ref{step:balls} assigns a label $\yh(B)$ to each ball $B$ in the focus region. The
label is one of $\{-1,+1,0\}$.  Labels $+1$ and $-1$ correspond to
a confident statement that $\eta_X(B)>0$ and $\eta_X(B)<0$
respectively; here $\eta_X(B)$ is the average value of $\eta(x)$ over points $x \in B \cap X$. The label $0$ indicates that the sign of $B$
is uncertain given the answers to the queries.

In step~\ref{step:examples} these labels are propagated to the
examples.  This is done in two sub-steps. First the {\em Possible
  Labels} for $x$, denoted $\PL_\ell(x)$, are calculated. $\PL_\ell(x)$
contains $+1$ if there is a ball containing $x$ whose label is $+1$,
symmetrically for $-1$. The label $\yh_\ell(x)$ is determined by the
possible labels set $\PL_\ell(x)$ as described in
Equation~\ref{eq:provisional-label}. The most significant label is ``!'', which
indicates balls with conflicting polarity containing $x$. Points
labeled ``!'' determine the uncertainty region and the focused query
region in the following level and the process repeats.

@@

The simplified algorithm addresses the first challenge of ``where
to query'' (see challenges in the introduction). It does not address
the other two challenges, which the detailed algorithm does address.

\paragraph{No smoothness assumptions.} As discussed in the introduction, one
is \emph{never} sure of having correctly determined the label of any 
specific point $x$. Thus in addition to focused (active) querying, we also
do background sampling of the entire space to pick up on possible mind
changes. For the sake of simplicity, we make one background query per focused query.

A related problem is that having the outer loop iterate over levels will
cause the algorithm to miss uncertain regions that later appear in lower
levels as a result of background sampling (the so-called mind changes).
Instead, each iteration of the full algorithm looks for the lowest level 
in which the uncertainty region is non-empty, and makes just two queries, 
one focused and one background.

\paragraph{Managing sampling.} As described in the introduction, we would
like to use queries that fall in the intersection of two balls for estimating
the bias in each of the balls. To facilitate query re-use, it is helpful to 
use Poisson sampling, as described in Section~\ref{sec:poisson}.

\iffalse
\begin{enumerate}
\item {\bf No smoothness assumptions:}\\
  As described in the introduction, one can \emph{never} be sure of having
  correctly determined the label of $x$. Thus in addition to focused
  (active) querying, we also do background sampling of the entire
  space to pick up on possible mind changes. For simplicity, we make
  one background query per focused query.

@@

\item {\bf Managing Sampling:}\\
  As described in the introduction, we would like to use queries that
  fall in the intersection of two balls for estimating the bias in
  both balls. However, selecting examples IID at random from the balls
  can create an undesired dependence between the estimates. This is solved using Poisson sampling, as described in section~\ref{sec:poisson}.
\end{enumerate}
\fi

@@

Once these queries are made, bias estimates $\widehat{\eta}(B)$ are
updated, the values of 
$\yh(B), \PL_\ell(x), \yh_\ell(x), U_\ell$ are updated,
 and the next iteration begins.

@@

The uncertainty region at level $\ell$ consists of the known-unknowns
from level $\ell-1$, that is, points $x \in X$ with
$\yh_{\ell-1}(x) = \, !$. This region is revealed to us piecemeal, a
few points at a time. To get good label complexity we need to actively
query the part of it we know. As a result, we are repeatedly in situations where the current region of uncertainty---in particular, the balls $B$ whose labels $\yh(B)$ we wish to determine---already contains some labels from earlier queries, and we wish to reuse those labels in a manner that is statistically sound. We manage this through \emph{Poisson
  sampling}~\cite{ghosh2002sampling} (Fig.~\ref{alg:sampling}).

As an example of what we are able to achieve, suppose we want to determine the label of a ball $B$ at level $\ell$ and have already queried points in several other balls, at levels $\leq \ell$, that intersect $B$. We will be able to apply \emph{all} these queries that fall in $B$ towards estimating $B$'s bias.

Here is the first key idea. In order to infer the label $\yh(B)$ of a ball $B$, we need to estimate the average bias of the ball, $\eta_X(B)$, within $\gamma$, which requires querying roughly $k = \tilde{O}(1/\gamma^2)$ random points in the ball. Instead of picking $k$ points at random in $B$, we query every point in $B$ with probability $\approx k/|X_B|$, \emph{independently}. More precisely, if $B$ lies at level $\ell$ then it has roughly $n/2^{\ell+1}$ points, so we query every point in $B$ with probability $\tau_\ell = \min(2^{\ell+2}k/n, 1)$.

The second key idea is to manage this independent sampling in a manner that makes it easy to reuse earlier queries.  At the outset of the algorithm, each instance $x \in X$ is assigned a random draw from the uniform distribution over $[0,1]$, denoted $T_x$. To query each point in a ball $B$ with probability $\tau_\ell$, we simply select all points in $x \in B$ that have $T_x \leq \tau_\ell$. If a point $x$ was queried earlier---because it also belongs to a ball $B'$ at level $\ell' \leq \ell$, say---then it will now get reused (because $T_x \leq \tau_{\ell'} \implies T_x \leq \tau_\ell$).

%Points are sampled in the order of increasing $T_x$ under some
%conditions described below.  Note that restricting the sampled points
%to be from set $A$ and having $T_x \leq \tau$ is the same, in
%expectation to sampling each point in $A$ independently at random with
%probability $\tau$.

%To infer the bias of a ball $B$, we need labels for
%$k = \tilde{O}(1/\gamma^2)$ random points in it. Balls at level $\ell$
%have roughly $n/2^{\ell+1}$ points, so we need about $2^{\ell+1} k/n$
%fraction of the ball to be queried. Instead, we query every point in
%the ball with probability $\tau_\ell = \min(2^{\ell+2}k/n, 1)$,
%%%\emph{independently}. Specifically, we query points $x$ in the ball
%with $T_x \leq \tau_\ell$, where each $x \in X$ is assigned a
%uniform-random value $T_x \in [0,1]$ at the outset of the algorithm.

Once a point is queried, it is never queried again; but see the note at the end of Section~\ref{sec:discrete} about situations in which it is permissible to query a point multiple times.

@@

Initialize uncertainty regions at all levels: 
\begin{itemize}
\item $U_0 = X$ and $U_\ell = \emptyset$ for $\ell \geq 1$
\end{itemize}
%\vspace{.1in}
